---
title: "R preprocessing - step 0"
output: html_notebook
---

# Load packages

here only RC9, save all for batch effect correction

```{r load packages, echo=FALSE}
## requirements
require(scater)
require(Matrix)
require(SingleCellExperiment)
require(scran)
library(TxDb.Mmusculus.UCSC.mm10.ensGene)
library(org.Mm.eg.db)
options(stringsAsFactors = FALSE)
require(RColorBrewer)
require(slingshot)
require(batchelor)
require(irlba)

## PsiNorm package
require(scone)
```

# Load Data

```{r load data, echo=FALSE}
matrix_dir = "/Users/cleliacorridori/Dropbox_2021\ Dropbox/Jorah\ Mormont/Code/Work/0_PreliminaryAnalysis_scRNA-seq/filtered_feature_bc_matrix/"
barcode.path <- paste0(matrix_dir, "barcodes.tsv.gz")
features.path <- paste0(matrix_dir, "features.tsv.gz")
matrix.path <- paste0(matrix_dir, "matrix.mtx.gz")

mat <- readMM(file = matrix.path)
feature.names = read.delim(features.path, 
                           header = FALSE,
                           stringsAsFactors = FALSE)
barcode.names = read.delim(barcode.path, 
                           header = FALSE,
                           stringsAsFactors = FALSE)
colnames(mat) = barcode.names$V1
rownames(mat) = feature.names$V1
#remove the -1 on cell IDs
colnames(mat) <-  sapply(strsplit(colnames(mat), split = "-", fixed=T), function(x) (x[1])) #strsslit makes list, we want 1. part as no list

```

# SCE Object

```{r}
sce <- SingleCellExperiment(assays = list(counts = mat))

dim(sce) # dataset dimension=genesxcells

rowData(sce)$ENSEMBL <- rownames(sce) #genes

# even if there are short names now, use old way to get positions
symb <- mapIds(org.Mm.eg.db, keys=rownames(sce), keytype="ENSEMBL", column="SYMBOL")
rowData(sce)$SYMBOL <- symb #other version of the genes name (the one that we use)

new.names <- rowData(sce)$SYMBOL
missing.name <- is.na(new.names)
new.names[missing.name] <- rowData(sce)$ENSEMBL[missing.name] #take the ID if there is no symbol
dup.name <- new.names %in% new.names[duplicated(new.names)]

new.names[dup.name] <- paste0(new.names, "_", rowData(sce)$ENSEMBL)[dup.name]
rownames(sce) <- new.names


tail(rownames(sce)) # row indices
tail(rowData(sce))  # data set

```

```{r}
# set channel using a reference one  (?) --> TxDb.Mmusculus.UCSC.mm10.ensGene
location <- mapIds(TxDb.Mmusculus.UCSC.mm10.ensGene, keys=rowData(sce)$ENSEMBL, 
    column="CDSCHROM", keytype="GENEID")
rowData(sce)$CHR <- location

summary(location=="chrM")

```

# Time information (hash info)

```{r}
hash_dir = "/Users/cleliacorridori/Dropbox_2021\ Dropbox/Jorah\ Mormont/Code/Work/input_data/"
hash.path <- paste0(hash_dir, "RC9_seurat_hashID.csv")
tags <- read.csv(hash.path)
head(tags)
# dim(tags) #for 16194 cells, not for alle the ones of the dataset (16233)
```

Check how many cells belong to each time point

```{r}
table(tags$hash.ID)
```

We use only the cells that have the time information

```{r}
sce <- sce[, colnames(sce) %in% tags$X]
dim(sce)
identical(colnames(sce), tags$X)
```

```{r}
sce$hashID <- tags$hash.ID

```


# Removing Double and Negative cells

```{r}
is.doublet <- sce$hashID == "Doublet"   
sce <- sce[, !is.doublet]    

is.neg <- sce$hashID == "Negative" 
sce <- sce[, !is.neg]
dim(sce)

```

# QC

```{r}
per.cell  <-perCellQCMetrics(sce, subsets=list(Mito=which(location=="chrM")), threshold=0)

summary(per.cell$total)
summary(per.cell$detected)
```

```{r, echo=FALSE}
colData(sce) <- cbind(colData(sce), per.cell)

plotColData(sce, x = "sum", y="subsets_Mito_percent", other_fields="hashID") + facet_wrap(~hashID)
```

outlier

```{r}
qc.stats <- quickPerCellQC(per.cell, percent_subsets="subsets_Mito_percent")

colSums(as.matrix(qc.stats))
```

```{r}
sce <- sce[,!qc.stats$discard]
plotColData(sce, x = "sum", y="subsets_Mito_percent", 
   other_fields="hashID") + facet_wrap(~hashID)
```



```{r}
#plotHighestExprs(sce, n=50) 
```

```{r}
ave <- calculateAverage(sce)
# Average GE per cell
rowData(sce)$AveCount <- ave
```


```{r}
# nexprs: Counting the number of non-zero counts in each row (per features) --> #cell per features
num.cells <- nexprs(sce, byrow = T)
smoothScatter(log10(ave), log(num.cells), ylab="Number of cells", 
    xlab=expression(Log[10]~"average count"))
```

## Remove genes

remove all that is not expressed in less 20 cells

```{r}
keep_feature <- num.cells > 20
sce <- sce[keep_feature,]  
table(keep_feature)
```

# Output data

All the printed/saved matrices have the first column with the gene names and the first row with the cell names. In the file name the signature "w00" means "with time 00h".

```{r}
cell_names <- matrix(colnames(sce)) #cells
tr_CellNames <- t(cell_names)
gene_names <- matrix(rownames(sce)) #genes

dim(cell_names)
dim(gene_names)

```

## LogNorm DATA

```{r}
# raw
raw_GE <- counts(sce)
raw_GE <- matrix(data = raw_GE, nrow = 13833  , ncol = 9894)
assay(sce, "counts", withDimnames=FALSE) <- raw_GE

# LogNorm
sce <- logNormCounts(sce)

# PsiNorm
sce<-PsiNorm(sce,  whichAssay = "counts")
sce<-logNormCounts(sce, name="psicounts")

assays(sce)$psicounts[1:3,1:3]
```

```{r save_logNorm}
Norm_GE <- logcounts(sce)
Norm_GE <- matrix(data = Norm_GE, nrow = 13833  , ncol = 9894)

Norm_GE.all <- rbind(tr_CellNames, Norm_GE)
dim(Norm_GE.all)

Norm_GE.all <- cbind(rbind(1,gene_names), Norm_GE.all)
dim(Norm_GE.all)

write.table(Norm_GE.all, "logNorm_check.csv",  sep = ",", col.names = NA)

```

# Select Informative Genes

modeling mean-variance trend follow new selection of features <https://osca.bioconductor.org/feature-selection.html#variance-of-the-log-counts>

```{r}
dec <- modelGeneVar(sce, block= sce$sample)
#Model the variance of the log-expression profiles for each gene, decomposing it into technical and biological components based on a fitted mean-variance trend.
```

```{r}
per.block <- dec$per.block
for (i in seq_along(per.block)) {
    decX <- per.block[[i]]
    plot(decX$mean, decX$total, xlab="Mean log-expression", 
        ylab="Variance", main=names(per.block)[i])
    curve(metadata(decX)$trend(x), col="blue", add=TRUE)
}
```


```{r}
# Get the top 2000 genes.
top.hvgs <- getTopHVGs(dec, n=2000)
str(top.hvgs)
```

```{r}
# Signatures from Carbognin et al. 2023
top.signs <- read.table("6signs.txt", header = TRUE, sep = "", dec = ".")
top.signs <- top.signs[,1]
str(top.signs)

all.genes <- rownames(sce)

present.signs <-all.genes[all.genes %in% top.signs]

```

# Checkpoint to save

```{r}
# save.image(file = "My_Object_check.RData")
```

```{r}
system("say Done!")
```

